Visual narratives of the Covid-19 pandemic

### 2022-01-17 JDSSV_manuscript_knitr.Rnw
update Jan 17

At the same time we still see a lot of defective graphics disseminated and shared: some that violate fundamental visualisation and statistical reporting principles such as accuracy, relevance, timeliness, clarity, coherence, and reproducibility. These principles have been laid out in numerous standards for statistical reporting in the application areas, such as the ESS standard for quality reporting, the CONSORT, PRISMA, CHEERs guidelines, and others (see https://equator-network.org). Numerous publications, initiatives, and ideas to improve the communication of quantitative and statistical information have been prepared, see for example \cite{Hoffrage2261,Tufte2001,Rosling2011,otavamylona2020}.

The Covid-19 pandemic has provided ample evidence that policy measures are only effectively implemented and adhered to by the people if they are accepted by a vast majority of the population. To achieve this goal, effective communication of quantitative evidence and statistical results among scientists, governments, media, and the citizens is essential to justify the adequacy, usefulness, and relevance of the measures.

A well proven method to effective communication lies in representing information in accessible ways (Gigerenzer et al., 2007; Gigerenzer and Edwards, 2011) by saying, for instance, “one in 10” instead of 10\%. Using absolute in place of relative numbers, representing information in appealing graphic forms, and summarizing most relevant facts in “fact boxes” are some of the methods developed and advocated for increasing transparency in communication between stakeholders such as health providers and patients (See, for example: https:// www. hardingcenter.de)
Fact boxes combined with icon arrays are recommended for the presentation of test results. Both representations are based on natural frequencies \citep{[62],Krauss2020} and present case numbers as simply and concretely as possible. Many scientific studies show that icon arrays help people understand numbers and risks more easily (e.g. \cite{McDowell2019}). The Harding Center for Risk Literacy shows many other examples of transparent communication of risks, including COVID-19\footnote{\url{https://www.hardingcenter.de/de/mrna-schutzimpfung-gegen-covid-19-fuer-aeltere-menschen}}.

Nevertheless, both the complexity of the phenomena and the "bipolarity" of statistical thinking remain challenges.

While human thinking tends towards pattern simplification and political communication also prefers a simple cause-effect relationship, real phenomena are often multivariate. Thus, when studying COVID-19 and predicting its spread, it is not only important to consider its symptomatology, the incidence and geographic distribution of diseases, population behavior patterns, government policies and impacts on the economy, on schools, on people in nursing homes and on social life as a whole, but to integrate these into the data analyses and the communication of results. Associations observed in the data can often be caused by third-party variables (confounders). In addition, much of the data comes from observational studies, which usually makes a robust causal attribution problematic. Statisticians calling out these limitations, however, face the danger that their statements might be pulled out one side in a polarized debate \citep{McConway2021}.

>>>>

%shall we include a list of newspapers/media outlets that we aim at covering?

At the same time we still see a lot of defective graphics disseminated and shared: some that violate fundamental visualisation and statistical reporting principles such as accuracy, relevance, timeliness, clarity, coherence, and reproducibility. These principles have been laid out in numerous standards for statistical reporting in the application areas, such as the ESS standard for quality reports \citep{ess2009}, the CONSORT, PRISMA, CHEERs guidelines, and others (see https://equator-network.org). Numerous publications, initiatives, and ideas to improve the communication of quantitative and statistical information have been prepared, see for example \cite{Hoffrage2261,Tufte2001,Rosling2011,otavamylona2020}.

The Covid 19 pandemic has amply demonstrated that policies are only effectively implemented and followed by the population if they are accepted by a large majority of the population. To achieve this goal, effective communication of quantitative evidence and statistical results between scientists, governments, the media and citizens is essential to justify the appropriateness, usefulness and relevance of the measures.

A best practice for effective communication is to present information in an understandable way (Gigerenzer et al., 2007; Gigerenzer and Edwards, 2011), for example by saying "one in ten" instead of 10\%. Using absolute rather than relative numbers, presenting information in an appealing graphic form and summarising the most important facts in "fact boxes" are some of the methods that have been developed and advocated to increase transparency in communication between stakeholders such as healthcare providers and patients (see for example: https:// www.hardingcenter.de).
Fact boxes in combination with icon arrays are recommended for the presentation of test results. Both representations are based on natural frequencies \citep{[62],Krauss2020} and represent case numbers as simply and concretely as possible. Many scientific studies show that icon arrays help people to understand numbers and risks more easily (e.g. \cite{McDowell2019}). The Harding Center for Risk Literacy shows many other examples of transparent communication of risks, including COVID-19\footnote{\url{https://www.hardingcenter.de/de/mrna-schutzimpfung-gegen-covid-19-fuer-aeltere-menschen}}.

Nevertheless, both the complexity of phenomena and the "bipolarity" of statistical thinking remain a challenge.

While human thinking tends to simplify patterns and political communication also prefers a simple cause-effect relationship, real phenomena are often multivariate. Thus, in studying COVID-19 and predicting its spread, it is not only important to consider symptomatology, disease incidence and geographic distribution, population behaviour patterns, government policies, and impacts on the economy, schools, people in nursing homes, and society at large, but also to incorporate these into data analyses and communication of results. Associations observed in the data can often be caused by third party variables (confounders). In addition, much of the data comes from observational studies, which usually makes robust causal attribution problematic. However, statisticians who point out these limitations are at risk of having their statements pulled out one-sidedly in a polarised debate \citep{McConway2021}.
